---
title: "Data Workshop with Marina 2026-03-06"
author: "Kari Inch"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---
## Writing Markdown Text

# Level 1 header

## Level 2 header (renders with a horizontal line underneath)

### Level 3 header

#### Level 4 header

**Bold text**

*Italic*

I can

-Make
-A
-List

I can also link [super cool websites](https://restorationfundorg)

**To write code** ctl+alt+i OR Code --> Insert Chunk

```{r}
2 + 2
```


```{r}
#| code-fold: true

# This code will only show up if you click on a dropwdown to open it
2 + 2
```

```{r}
#| echo: false

# Your code will not show at all, only the output
2 + 2
```

```{r}
#| include: false

# This code will run, but neither the code nor the output will render
2 + 2
```

# Penguins Analysis

## About

Purpose of this script 

omthing about the data and anything important that you would need to remember


## Setup

```{r}
# Load libraries ----
library(tidyverse)
library(here)
library(janitor)
library(lubridate) # date and time wrangling
library(gghighlight)
library(sysfonts)
library(showtext)

# Read in data ----
penguins_raw <- read_csv(here("data", "raw", "messy_penguins.csv"))
nesting_raw <- read_csv(here("data", "raw", "nesting.csv"))
```


## Data exploration

```{r}
# What columns are in the penguins data?
colnames(penguins_raw)

# How many penguin species
n_distinct(penguins_raw$Species) # use $ to call a specific column from a dataframe

# What are they?
unique(penguins_raw$Species)

# What islands?
unique(penguins_raw$island)
```

## Data cleaning
```{r}
# Peek at the data
head(penguins_raw) # prints out a snip-it of the dataframe in the console
```


```{r}
# Clean up column names
penguins_clean <- penguins_raw %>% 
  clean_names()

# Check columns fix
colnames(penguins_clean)
```

```{r}
# Island column ----

# Fix inconsistent capitalization in the column
penguins_clean <- penguins_clean %>% 
  mutate(island = str_to_title(island))

# Check
unique(penguins_clean$island)
```

```{r}
# Bill length column ----

range(penguins_clean$bill_length_mm, na.rm = TRUE) #means to look at non NA values

class(penguins_clean$bill_length_mm)

# What values do I have that are not recognized as numeric
penguins_clean %>% 
  count(bill_length_mm) %>% 
  filter(is.na(as.numeric(bill_length_mm)))

penguins_clean <- penguins_clean %>% 
  mutate(bill_length_mm = as.numeric(bill_length_mm))

# Check range
range(penguins_clean$bill_length_mm, na.rm = TRUE) # Oops can't have negative numbers...let's fix that!

penguins_clean <- penguins_clean %>% 
  mutate(bill_length_mm = case_when(bill_length_mm == -999.0 ~ NA,
                                    .default = bill_length_mm))

# Check range
range(penguins_clean$bill_length_mm, na.rm = TRUE)
```

```{r}
# Observation date column ----
head(penguins_clean$observation_date)

# Fix datetime
penguins_clean <- penguins_clean %>% 
  mutate(date = parse_date_time(observation_date, 
                                orders = c("dmy", "ymd", "mdy",
                                           "ymd HMS"))) #make a new column to work from and then delete the original column

head(penguins_clean$date)

# Do not need time
penguins_clean <- penguins_clean %>% 
  separate(col = date,
           into = c("date", "time"),
           sep = " ") %>% 
  mutate(date = as.Date(date))

# Drop columns we don't need after we check and are happy with the above output
penguins_clean <- penguins_clean %>% 
  select(-c(time, observation_date))

# Create some new variables based on date
penguins_clean <- penguins_clean %>% 
  mutate(year = year(date), 
         month = month(date, label = TRUE, abbr = FALSE),
         day = day(date),
         day_of_year = yday(date))
```

```{r}
# Check for dublicates
penguins_clean %>% 
  get_dupes() # Now is the time to figure out what to do with the duplicates and how you want to proceed with them, check with data enter-er, etc...
```

```{r}
# Remove duplicates (I talked with the field team!)
# Count rows before removing
nrow(penguins_clean)

#Remove dupes
penguins_clean <- penguins_clean %>% 
  distinct()

# Count after removing
nrow(penguins_clean)
```

```{r}
# Save cleaned penguin data ----
write_csv(penguins_clean, here("data", "processed", "penguins.csv"))
```


## Visualization

```{r}
# Simple bar chart with What, Where, and When

pen_bar <- ggplot(penguins_clean %>% 
         mutate(species = fct_rev(fct_infreq(species)))) +
  geom_bar(aes(x = species)) +
  geom_text(aes(x = species, label = after_stat(count)),
            stat = "count",#puts numbers at the end of the columns
            hjust = 1.5,
            color = "white",
            size = 5) + 
  labs(title = "Penguin Counts at Palmer Station LTER",
       subtitle = "2007 - 2009",
       caption = "Data: Gorman et al. 2014 via {palmerpenguins}") +
  coord_flip() + #exception with journal articles...long labels remain on the x-axis...this is circumstantial 
  theme_minimal(base_size = 14) + #adjusts all text in plot together 
  theme(
    axis.title = element_blank(),
    axis.text.x = element_blank(),
    panel.grid = element_blank(),
    plot.caption = element_text(face = "italic", 
                                color = "grey50",
                                size = rel(0.7)), #this is a percentage of the base size
    plot.title.position = "plot",
    plot.title = element_text(face = "bold"),
    plot.margin = margin(20, 20, 20, 20))
```

```{r}
# Save penguins bar chart #to uncomment (#) we use ctrl+shift+c
# ggsave(here("figures", "pen_bar.png"),
#        pen_bar, width = 6, height = 4) #start with dimensions like this and then work from there
```

```{r}
# Penguin size scatterplot 
ggplot(data = penguins_clean) + #I could put the "aes" into this area which is global if I wanted everything to have the same aes
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm,
                 color = species, 
                 shape = species),
             size = 3,
             alpha = 0.8) + #this allows transparancy of the colors
  geomtextpath::geom_labelsmooth(aes(label = species,
                                     x = bill_length_mm,
                                     y = bill_depth_mm, 
                                     color = species),
                                 method = "lm", 
                                 size = 5,
                                 linewidth = 1) +
  # geom_smooth(aes(x = bill_length_mm,
  #                 y = bill_depth_mm,
  #                 color = species), #when you add in groups this makes the trend line opposite of what it would be if you didn't --> Simpson's Paradox
  #             method = "lm", se = FALSE) +
  scale_color_manual(values = c("darkorange", "purple",
                                "cyan4")) +
  labs(title = "Penguin Bill Dimensions",
       subtitle = "Bill Length & Depth for Adelie, Chinstrap, and Gentoo Penguins at Palmer Station LTER",
       x = "Bill Length (mm)",
       y = " Bill Depth (mm)",
       color = "Penguin Species",
       shape = "Penguin Species") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    plot.margin = margin(10, 10, 10, 10),
    axis.title.x = element_text(margin = margin(15, 0, 0, 0)), #increase margins around axis titles
    axis.title.y = element_text(margin = margin( 0 , 15, 0, 0)),
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```





















