---
title: "Data Workshop with Marina 2026-03-06"
author: "Kari Inch"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---
## Writing Markdown Text

# Level 1 header

## Level 2 header (renders with a horizontal line underneath)

### Level 3 header

#### Level 4 header

**Bold text**

*Italic*

I can

-Make
-A
-List

I can also link [super cool websites](https://restorationfundorg)

**To write code** ctl+alt+i OR Code --> Insert Chunk

```{r}
2 + 2
```


```{r}
#| code-fold: true

# This code will only show up if you click on a dropwdown to open it
2 + 2
```

```{r}
#| echo: false

# Your code will not show at all, only the output
2 + 2
```

```{r}
#| include: false

# This code will run, but neither the code nor the output will render
2 + 2
```

# Penguins Analysis

## About

Purpose of this script 

omthing about the data and anything important that you would need to remember


## Setup

```{r}
# Load libraries ----
library(tidyverse)
library(here)
library(janitor)
library(lubridate) # date and time wrangling
library(gghighlight)
library(sysfonts)
library(showtext)

# Read in data ----
penguins_raw <- read_csv(here("data", "raw", "messy_penguins.csv"))
nesting_raw <- read_csv(here("data", "raw", "nesting.csv"))
```


































