---
title: "Data Workshop with Marina 2026-03-06"
author: "Kari Inch"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---
## Writing Markdown Text

# Level 1 header

## Level 2 header (renders with a horizontal line underneath)

### Level 3 header

#### Level 4 header

**Bold text**

*Italic*

I can

-Make
-A
-List

I can also link [super cool websites](https://restorationfundorg)

**To write code** ctl+alt+i OR Code --> Insert Chunk

```{r}
2 + 2
```


```{r}
#| code-fold: true

# This code will only show up if you click on a dropwdown to open it
2 + 2
```

```{r}
#| echo: false

# Your code will not show at all, only the output
2 + 2
```

```{r}
#| include: false

# This code will run, but neither the code nor the output will render
2 + 2
```

# Penguins Analysis

## About

Purpose of this script 

omthing about the data and anything important that you would need to remember


## Setup

```{r}
# Load libraries ----
library(tidyverse)
library(here)
library(janitor)
library(lubridate) # date and time wrangling
library(gghighlight)
library(sysfonts)
library(showtext)

# Read in data ----
penguins_raw <- read_csv(here("data", "raw", "messy_penguins.csv"))
nesting_raw <- read_csv(here("data", "raw", "nesting.csv"))
```


## Data exploration

```{r}
# What columns are in the penguins data?
colnames(penguins_raw)

# How many penguin species
n_distinct(penguins_raw$Species) # use $ to call a specific column from a dataframe

# What are they?
unique(penguins_raw$Species)

# What islands?
unique(penguins_raw$island)
```

## Data cleaning
```{r}
# Peek at the data
head(penguins_raw) # prints out a snip-it of the dataframe in the console
```


```{r}
# Clean up column names
penguins_clean <- penguins_raw %>% 
  clean_names()

# Check columns fix
colnames(penguins_clean)
```

```{r}
# Island column ----

# Fix inconsistent capitalization in the column
penguins_clean <- penguins_clean %>% 
  mutate(island = str_to_title(island))

# Check
unique(penguins_clean$island)
```

```{r}
# Bill length column ----

range(penguins_clean$bill_length_mm, na.rm = TRUE) #means to look at non NA values

class(penguins_clean$bill_length_mm)

# What values do I have that are not recognized as numeric
penguins_clean %>% 
  count(bill_length_mm) %>% 
  filter(is.na(as.numeric(bill_length_mm)))

penguins_clean <- penguins_clean %>% 
  mutate(bill_length_mm = as.numeric(bill_length_mm))

# Check range
range(penguins_clean$bill_length_mm, na.rm = TRUE) # Oops can't have negative numbers...let's fix that!

penguins_clean <- penguins_clean %>% 
  mutate(bill_length_mm = case_when(bill_length_mm == -999.0 ~ NA,
                                    .default = bill_length_mm))

# Check range
range(penguins_clean$bill_length_mm, na.rm = TRUE)
```

```{r}
# Observation date column ----
head(penguins_clean$observation_date)

# Fix datetime
penguins_clean <- penguins_clean %>% 
  mutate(date = parse_date_time(observation_date, 
                                orders = c("dmy", "ymd", "mdy",
                                           "ymd HMS"))) #make a new column to work from and then delete the original column

head(penguins_clean$date)

# Do not need time
penguins_clean <- penguins_clean %>% 
  separate(col = date,
           into = c("date", "time"),
           sep = " ") %>% 
  mutate(date = as.Date(date))

# Drop columns we don't need after we check and are happy with the above output
penguins_clean <- penguins_clean %>% 
  select(-c(time, observation_date))

# Create some new variables based on date
penguins_clean <- penguins_clean %>% 
  mutate(year = year(date), 
         month = month(date, label = TRUE, abbr = FALSE),
         day = day(date),
         day_of_year = yday(date))
```

```{r}
# Check for dublicates
penguins_clean %>% 
  get_dupes() # Now is the time to figure out what to do with the duplicates and how you want to proceed with them, check with data enter-er, etc...
```

```{r}
# Remove duplicates (I talked with the field team!)
# Count rows before removing
nrow(penguins_clean)

#Remove dupes
penguins_clean <- penguins_clean %>% 
  distinct()

# Count after removing
nrow(penguins_clean)
```

```{r}

```
























